Exploratory data structure comparisons: Three new visual tools based on Principal Component Analysis

### 2018-06-18 article/Old versions/MPS/PCADSC_article.tex
Data analysis updated

\hl{However, there is a generic situation where the angles between the arrows lack discriminative power, namely when two or more neighbouring eigenvalues are identical or close. For instance, multiple eigenvalues close to 1 arise if more than one of the variables are uncorrelated, or close to uncorrelated, of all the other variables. In this situation only the corresponding eigenspace is well estimates, but the rotation of the associated eigenvectors will be subject to sampling variability. To investigate if we are in such a situation of identical data structures with neighbouring eigenvalues being close, where comparably long arrows may occur in the corresponding blocks of off-diagonal cells, we have implemented a permutation test similar as done for the CE-plot. For each random reallocation we recompute the angles $\theta_{jk}$. The resampled angles may both be used for quantitative tests, where large angles are critical for the diagonal cells and small angles are critical for the off-diagonal cells, as well as for the visualisation of the sampled angle regions. In the visualisation we propose to present the entire sampled region of angles for each cell, but grey shaded according to percentiles such that more critical angles are lighter grey. In our implementation we do not present the p-values for the quantitative tests, but the p-values are available in the R-object to support the visualization if necessary. However, it is important to remember that there are $d \times d$ p-values, which means that attention should be given to a possible multiple testing issue.}

@@

\hl{Figure \ref{plot.simAngle} presents two examples of angle plots based on simulated data. For dataset $\mathbf{A}$, we find very short off-diagonal arrows except for the off-diagonal cells corresponding to the 3'rd and 4'th eigenvalues. However, we see that the arrows in cell (3,4) and (4,3) lie well within the shaded angle regions (with associated p-values 0.9119 and 0.9103). This indicates that we are in a situation with identical data structure, and with 3'rd and 4'th eigenvalues being close (this corresponds to the two independent variables in \emph{model 1}). For dataset $\mathbf{B}$, on the other hand, we find very little agreement in all components, which should be expected, as the two groups in the dataset correspond to different data structures.}

@@

In the rankings of happiness, not much work has yet been devoted to evaluating the assumption of international comparabiltity, though Veenhoven presents a theoretically thorough, but empirically simplistic, summary of possible reasons for lack of comparability \cite{Veenhoven2012}, while Lolle \& Andersen show highly potent translation issues for the term \textit{happiness} \cite{Lolle2016}.

If for instance two countries differ in terms of how social networks are typically built and structured, with one emphasizing family relations and the other mostly focusing on other social relations, then having a weak family connection does not have the same implications in the first country as it does in the second one. More specifically, whereas in the first country, lack of familial network might be related to loneliness, lack of general social capital, and isolation, in the second country, the quality of the family network might not be informative at all about other aspects of a person's social or psychological well-being. The two countries thus differ in how different aspects or measures of psychological well-being are interrelated, which is essentially a difference in data structures. And therefore, comparing the two countries in these measures is not a meaningful endeavor.

@@

Moving on to the angle plot, we find that the differences are primarily to be found in the second, third, and fourth principal components (PCs). The blue arrows visualize the decomposition of the principal components for the Bulgarian dataset in the coordinate system of the Danish dataset. We see that PC2 also loads on PC3, that PC3 also loads on PC4, and that PC4 also loads on PC2 and PC3. The red arrows visualize the decomposition of the principal components for the Danish dataset in the coordinate system of the Bulgarian dataset. Here, we see that PC2 also loads on PC4, that PC3 also loads on PC2 and PC4, and that PC4 also loads on PC3. Thus, if we wish to understand why differences in the data structures occur, an inspection of the loadings of components 2, 3 and 4 might be informative.

@@

Though the PCADSC toolbox comprises three diagnostic plots as of now, the PCA decomposition can of course be utilized further to bring about new PCADSC plots that illustrate more aspects of data structure differences. For instance, the scree plot (as introduced by Cattell \cite{Cattell1966}), which plots the ordered eigenvalues against their component numbers, could easily be recreated as a PCADSC tool by creating scree plots for both data subsets in the same coordinate system, yielding a simple graphical comparison. The resampling approach used in the CE plot could then be adopted, thus annotating the scree plots with resampled curves. This general approach to model-fit evaluation is very helpful in combination with permutation tests, as inspired by Lin and colleagues \cite{LinEtAl2002}. Similarly, the angle plot could also be annotated with $p$-values from a permutation test based on appropriate test statistics.

>>>>

\hl{However, there is a generic situation where the angles between the arrows lack discriminative power, namely when two or more neighbouring eigenvalues are identical or close. For instance, multiple eigenvalues close to 1 arise if more than one of the variables are uncorrelated, or close to uncorrelated, of all the other variables. In this situation only the corresponding eigenspace is well estimates, but the rotation of the associated eigenvectors will be subject to sampling variability. To investigate if we are in such a situation of identical data structures with neighbouring eigenvalues being close, where comparably long arrows may occur in the corresponding blocks of off-diagonal cells, we have implemented a permutation test similar as done for the CE-plot. For each random reallocation we recompute the angles $\theta_{jk}$. The resampled angles may both be used for quantitative tests, where large angles are critical for the diagonal cells and small angles are critical for the off-diagonal cells, as well as for the visualisation of the sampled angle regions. In the visualisation we propose to present the entire sampled region of angles for each cell, but grey shaded according to percentiles such that more critical angles are lighter grey. In our implementation we do not present the p-values for the quantitative tests, but the p-values are available in the R-object to support the visualization if necessary. It is important to remember that there are $d \times d$ p-values, which means that attention should be given to a possible multiple testing issue.}

@@

\hl{Figure \ref{plot.simAngle} presents two examples of angle plots based on simulated data. For dataset $\mathbf{A}$, we find very short off-diagonal arrows except for the off-diagonal cells corresponding to the 3'rd and 4'th eigenvalues. But since the arrows in cells (3,4) and (4,3) lie well within the shaded angle regions (supported by p-values 0.9119 and 0.9103) this is in agreement with identical data structures, where the 3'rd and 4'th eigenvalues are close (in this situation corresponding to the two independent variables in \emph{model 1}). For dataset $\mathbf{B}$, on the other hand, we find very little agreement in all components, which should be expected, as the two groups in the dataset correspond to different data structures.}

@@

In the rankings of happiness, not much work has yet been devoted to evaluating the assumption of international comparability, though Veenhoven presents a theoretically thorough, but empirically simplistic, summary of possible reasons for lack of comparability \cite{Veenhoven2012}, while Lolle \& Andersen show highly potent translation issues for the term \textit{happiness} \cite{Lolle2016}.

If for instance two countries differ in terms of how social networks are typically built and structured, with one emphasizing family relations and the other mostly focusing on other social relations, then having a weak family connection does not have the same implications in the first country as it does in the second one. More specifically, whereas in the first country, lack of familial network might be related to loneliness, lack of general social capital, and isolation, in the second country, the quality of the family network might not be informative at all about other aspects of a person's social or psychological well-being. The two countries thus differ in how different aspects or measures of psychological well-being are interrelated, which is essentially a difference in data structures. And therefore, comparing the two countries in these measures is not a meaningful endeavour.

@@

Moving on to the angle plot, we find that the differences are primarily to be found in the second, third, and fourth principal components (PCs). The blue arrows visualize the decomposition of the principal components for the Bulgarian dataset in the coordinate system of the Danish dataset. We see that PC2 also loads on PC3, that PC3 also loads on PC4, and that PC4 also loads on PC2 and PC3. The red arrows visualize the decomposition of the principal components for the Danish dataset in the coordinate system of the Bulgarian dataset. Here, we see that PC2 also loads on PC4, that PC3 also loads on PC2 and PC4, and that PC4 also loads on PC3. \hl{These statements are supported by p-values for the hypotheses of no loading that all are less than $0.002$.} Thus, if we wish to understand why differences in the data structures occur, an inspection of the loadings of components 2, 3 and 4 might be informative.

@@

Though the PCADSC toolbox comprises three diagnostic plots as of now, the PCA decomposition can of course be utilized further to bring about new PCADSC plots that illustrate more aspects of data structure differences. For instance, the scree plot (as introduced by Cattell \cite{Cattell1966}), which plots the ordered eigenvalues against their component numbers, could easily be recreated as a PCADSC tool by creating scree plots for both data subsets in the same coordinate system, yielding a simple graphical comparison. The resampling approach used in the CE plot could then be adopted, thus annotating the scree plots with resampled curves. This general approach to model-fit evaluation is very helpful in combination with permutation tests, as inspired by Lin and colleagues \cite{LinEtAl2002}.

%Similarly, the angle plot could also be annotated with $p$-values from a permutation test based on appropriate test statistics.
